---
title: "CodeBook"
output: 
  html_document:
    theme: united
    toc: yes
---

## Assignment Description, Coursera Getting and Cleaning Data - Course Progject: 
### Tasks 
* Merges the training and the test sets to create one data set.  
* Extracts only the measurements on the mean and standard deviation for each measurement.  
* Uses descriptive activity names to name the activities in the data set.  
* Appropriately labels the data set with descriptive variable names.  
* From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.  

### Data    
 https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip 


## My approach  
 1. Get the Data  
 1.0. Dowload files  
 1.1. Read into R  
 2. Explore the Data  
 2.0. Understand the structure of the data file(s)  
 2.1. check for N/A values (blanks)  
 2.2. verify unique variable names  
 2.3. review profile (specifically are there any outliers)  
 3. Shape the Data  
 3.0  combine datasets  
 3.1  extract only measurements on mean and standard deviation  
 3.2  free memory  
 4. Analyze the Data  


## 1. Get the Data 

### 1.0. Dowload files
```{r eval=FALSE, echo=TRUE}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
destfile <- "UCI_HAR.zip"
download.file(url, destfile, method='curl')
```

Create folder, extract the archived files, and write them to the newly created folder:
```{r eval=FALSE, echo=TRUE}
unzip(destfile)
```

Show which files have been extracted:
```{r eval=TRUE, echo=TRUE}
files <- list.files("UCI HAR Dataset")
```
```{r eval=TRUE, echo=FALSE}
files
```
Note that "test" and "train" are folders.
 
Explore folders:
```{r eval=TRUE, echo=TRUE}
testFiles <- list.files("UCI HAR Dataset/test")
```
```{r eval=TRUE, echo=FALSE}
testFiles
```
And subfolder:
```{r eval=TRUE, echo=TRUE}
testInertialSignalsFiles <- list.files("UCI HAR Dataset/test/Inertial Signals")
```
```{r eval=TRUE, echo=FALSE}
testInertialSignalsFiles
```

The train folder has a similar structure.

### 1.1. Read into R
```{r eval=TRUE, echo=TRUE}
trn_j <- read.table("UCI HAR Dataset/train/subject_train.txt")
tst_j <- read.table("UCI HAR Dataset/test/subject_test.txt")
trn <- read.table("UCI HAR Dataset/train/X_train.txt")
tst <- read.table("UCI HAR Dataset/test/X_test.txt")
trn_b <- read.table("UCI HAR Dataset/train/y_train.txt")
tst_b <- read.table("UCI HAR Dataset/test/y_test.txt")
act_b <- read.table("UCI HAR Dataset/activity_labels.txt")
feat <- read.table("UCI HAR Dataset/features.txt")
```

## 2. Explore the Data

### 2.0. Understand the structure of the data file(s)
```{r eval=TRUE, echo=TRUE}
# trn[] and tst[] should have similar structures
if(length(trn_j) == length(tst_j))  exp_length_j = 1 
if(length(trn) == length(tst))  exp_length = 1 
if(length(trn_b) == length(tst_b))  exp_length_b = 1 
if(!exp_length_j || !exp_length || !exp_length_b) warning("Input files are not similar.")
```
### 2.1. check for N/A values (blanks)
```{r eval=TRUE, echo=TRUE}
if(length(trn_j[is.na(trn_j[ , 1])]) != 0) warning("The file subject_train.txt included blanks")
if(length(tst_j[is.na(tst_j[ , 1])]) != 0) warning("The file subject_test.txt included blanks")
if(length(trn[is.na(trn[ , 1])]) != 0) warning("The file X_train.txt included blanks")
if(length(tst[is.na(tst[ , 1])]) != 0) warning("The file X_train.txt included blanks")
if(length(trn_b[is.na(trn_b[ , 1])]) != 0) warning("The file y_train.txt included blanks")
if(length(tst_b[is.na(tst_b[ , 1])]) != 0) warning("The file y_train.txt included blanks")
```

### 2.2. verify unique variable names
If they are not unique, append the an index to the name to distinquish it from others.
```{r eval=TRUE, echo=TRUE}
trn_names <- names(trn)
if(length(trn_names) > length(unique(trn_names))) warning("There are duplicate variable names")
tst_names <- names(tst)
if(length(tst_names) > length(unique(tst_names))) warning("There are duplicate variable names")
compare_names <- trn_names != tst_names
if(length(compare_names[FALSE]) != 0) warning("The column names are not the same in the data sets")
if(length(act_b[,2]) > length(unique(act_b[,2]))) 
  {
    activities <- as.data.frame(paste(act_b[,1], act_b[,2], sep='_'))
  } else { 
    activities <- as.data.frame(act_b[,2])
  }
if(length(feat[,2]) > length(unique(feat[,2]))) 
  {
    var_names <- as.data.frame(paste(feat[,2], feat[,1], sep='_'))
  } else { 
    var_names <- as.data.frame(feat[,2])
  }
names(var_names) <- c("Variable" )
```

### 2.3. review profile (specifically are there any outliers)
```{r eval=TRUE, echo=TRUE}
outliers <- function(x) {
  has_outliers = 0
  qnt_x <- quantile(x)
  iqr_x <- IQR(x)
  lower <- qnt_x["25%"] - (iqr_x*1.5)
  upper <- qnt_x["75%"] + (iqr_x*1.5)
  if((qnt_x["0%"] < lower) || (qnt_x["100%"] > upper )) has_outliers = 1
  has_outliers
}
```

This may not be critical to this particular analysis, but generally it can be a useful tool if there are issues with the analysis.
```{r eval=TRUE, echo=TRUE}
outlier_list <- apply(trn, 2, outliers)
```

## 3. Shape the Data  

### 3.0. Combine datasets
```{r eval=TRUE, echo=TRUE}
trn_tst_j <- rbind(trn_j, tst_j)
  names(trn_tst_j) <- c("Subject")
trn_tst_b <-rbind(trn_b, tst_b)
  names(trn_tst_b) <- c("Activity")
trn_tst <- rbind(trn, tst)
  names(trn_tst) <- var_names$Variable
```

### 3.1 Extract only the measurements on the mean and standard deviation
```{r eval=TRUE, echo=TRUE}
library("dplyr", lib.loc="/Library/Frameworks/R.framework/Versions/3.1/Resources/library")
trn_tst_mean <- select(trn_tst, contains("[Mm]ean|str"))
trn_tst_sdev <- select(trn_tst, contains("[Ss][Tt][Dd]|str"))
```
Create a single data frame with these two and capture names for use later.
```{r eval=TRUE, echo=TRUE}
trn_tst_msr <- cbind(trn_tst_mean, trn_tst_sdev)
msr_names <- names(trn_tst_msr)
```

Create One Dataset
```{r eval=TRUE, echo=TRUE}
trn_tst_all <-cbind(trn_tst_j, trn_tst_b, trn_tst_msr)
```

### 3.2 Free Memory
```{r eval=TRUE, echo=TRUE}
rm(trn_j, tst_j, trn, tst, trn_b, tst_b, act_b, feat)
rm(trn_tst_j, trn_tst_b, trn_tst_mean, trn_tst_sdev, trn_tst_msr)
```

## 4. Analyze the Data
```{r eval=FALSE, echo=TRUE}
getMeans <- function(Subject, Activity) {
  is_relevant <- trn_tst_all[,Subject==Subject & Activity==Activity]
  output <- as.data.frame(apply(trn_tst_all[is_relevant], 2, mean))
  output[1,] <- Subject
  if(Activity==1) output[2,] <- 'WALKING'
  if(Activity==2) output[2,] <- 'WALKING_UPSTAIRS'
  if(Activity==3) output[2,] <- 'WALKING_DOWNSTAIRS'
  if(Activity==4) output[2,] <- 'SITTING'
  if(Activity==5) output[2,] <- 'STANDING'
  if(Activity==6) output[2,] <- 'LAYING'  
  names(output) <- c(paste('Subject:', Subject, '_Activity:', output[2,], sep=''))
  output
}
rs_names <- c('Subject', 'Activity', msr_names)
resultSet <- data.frame(row.names=rs_names)
  for(Subject in 1:30) {
    for(Activity in 1:6) {
      resultSet <- cbind(resultSet, getMeans(Subject, Activity))
    }
  }
write.csv(resultSet, 'resultSet.csv')
```
